---
title: "Homework 2 - Priors and Data"
author: "Courtney Hodge"
date: "2024-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
wine <- read.csv("C:\\Users\\hodge\\Downloads\\whitewine-training-ds6040.csv")
```

# Part 1: Using Conjucate Priors with White-Wine Data (33pts)
* Choose 2 continuous predictors and plot density plots (using ggplot2)
```{r}
library(ggplot2)

ggplot(wine, aes(x = pH)) +
  geom_density(fill = "orange", alpha = 0.5) +
  ggtitle(paste("Density Plot of pH in wine data" ))

ggplot(wine, aes(x = alcohol)) + 
  geom_density(fill = "darkblue", alpha = 0.5) +
  ggtitle(paste("Density Plot of alcohol in wine data"))
```
* For each predictor, consider the density plot. What do you notice? Do they look like they
are normally distributed? Skewed?

> For the pH predictor, the denisty plot looks similar to a normal distribution centered around 0 pH with an extremely slight skew right, which is not bad at all. There's also symetry with this distribution. 
> For the alcohol predictor, there is not a normal distrbution of this data. If anything, it is very strongly skewed right, with a peak around -1.

## 1. 

Using a normal likelihood with known variance (the known variance for each variable is
the observed variance, go ahead and calculate that), 

use the appropriate conjugate prior
and calculate out the posterior distribution (you should be able to look up the formula for
the posteriors, no need to derive them out yourself). When you calculate these posterior
distributions, use two sets of hyperparameters (per variable), one where the
hyperparameters specify a fairly uninformative prior, and the other where the
hyperparameters are much more informative (this doesn't need to be a reasonable value
either, this exercise is to demonstrate the impact of hyperparameter choice). At the end of
this exercise, you should have the parameters for 4 posterior distributions.

```{r}
#calculate the sample variance and mean
mean_pH <- mean(wine$pH)
var_pH <- var(wine$pH)

mean_alcohol <- mean(wine$alcohol)
var_alcohol <- var(wine$alcohol)

#print results
print("The mean and variance of pH:")
mean_pH
var_pH

print("The mean and variance of alcohol:")

mean_alcohol
var_alcohol
```
Likelihood:
* The data is normally distributed w/ a known var of 1
$X_i$ ~ $$

Posterior for pH
* Uninformative Prior $(\mu_0 = 0, \sigma^2 = 1000):$
```{r}

```

* Informative Prior $(\mu_0 = 0.-2, \sigma^2 = 0.01):$
```{r}

```

